Species-specific effects of the Urban Heat Island on tree growth across Berlin

### 2020-03-07 analysis/paper/paper.Rmd
updated intro

pgbreak <- function() {
  if (grepl("docx", out_format)) {
    return("##### pagebreak")
  } else if (is_latex_output()) {
    return("\\newpage")
  }
}

@@

Berlin is subject to the most intense Urban Heat Island in Germany due to its large extent and development intensity [@kuttler2015].
Here, substantially higher temperatures have been observed for urban as compared to rural areas, with differences of up to 12 K during day-time and 6 K on average for night-time temperatures [2001-2010, @fenner2014].

Consequently, urban green (infrastructure) systems are  subjected to increased heat more frequently than rural ecosystems, potentially affecting  their process dynamics positively or adversely. 
Their performance and health, however, is closely tied to local energy budgets [@grimmond1996 ; @hertel2019], which in turn are decisive for controlling human wellbeing [e.g. @maras2016], amongst other factors.
Assessing the effect of increased temperatures on green infrastructure, as part of the urban landscape, is therefore instrumental for understanding, and ultimately mitigating, the potential impact of future warming and (continued) development  on increasingly urban societies [@norton2015 ].

Trees, in particular, provide shading as well as transpirative cooling in their vicinity [@gillner2015;@oke1982], and therefore can reduce ambient temperatures, infrastructure power-consumption and (human) thermal discomfort  [e.g. @gulyas2006; @akbari2001; @mayer1987; @hoyano1988]; simultaneously, they provide numerous other environmental, cultural and psychological services and/or benefits [see @tzoulas2007 for review].

Tree growth dynamics in urban areas are modulated by air temperature, amongst other factors, with a tendency for enhanced growth rates (compared to rural individuals) as reported by @zhao2016 and @pretzsch2017. 
Hence,  temperature increases may also allow using recent tree growth dynamics and trajectories as a proxy for on-going and future warming, providing an additional line of evidence to support the growing knowledge base on future climate- and human-vegetation and dynamics [@zhao2016].
For Berlin, @dahlhausen2018, showed divergent patterns for young and old trees dependent on urban development intensity for *Tilia cordata* Mill, the most abundant tree of the city, but overall reported an increased growth rate which they attributed to the urban heat island effect (UHI).
Similarly, @moser-reischl2019a showed positive associations between air temperature and radial growth for two species commonly selected by urban planners (*T. cordata*, *Rubinia pseudoacacia*) in Munich.
By contrast, @gillner2014 highlight decreased growth for *Acer* species  (*A. platanoides* and *pseudoplatanus*), *Platanus x hispanica* and *Quercus rubra* with higher summer temperatures of the preceding year, especially when compounded with drought, in another German metropolis (Dresden). 
**In the future.. drought, both soil moisture and atmospheric drought.. potential issue for street trees .. [@brune2016]
??**
[@zhao2016] and [@jia2018] direction.....

These results indicate a high degree of complexity in effect sizes and directions specific to species and locality.
Tree growth in urban settings is affected by other processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc. [@rhoades1999; @quigley2004; @randrup2001; @pauleit2002].
While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e. genera and/or species, and space).
This typically hinders the extrapolation from individual sampling sites toward prediction of effect sizes across entire urban areas and tree stocks. 
<!-- Therefore, studies reliant on labor-intensive sampling, must carefully address confounding factors to ensure effect sizes are estimated adequately. -->

To complement detailed dendroecological analyses in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individual trees provided by the Berlin Senate Administration (Senatsverwaltung). 
This data set contains information on location, species, trunk diameter (at breast height; DBH), and height, amongst other variables. 
In a space-for-time substitution, the absolute growth potential of species can be assessed across the entire urban area, and related to effects of the UHI, while accounting for other location-specific factors, such as street characteristics, development intensity, available soil volume, etc.
[@quigley2004] .. inferred absolute growth potential for species across demographic strata

@@

# Methods

# Results

@@

# Discussion

# Conclusion

@@

<!-- The following line inserts a page break  -->
\newpage

@@

\newpage

>>>>

out_format <- knitr::opts_knit$get("rmarkdown.pandoc.to")

pgbreak <- function() {
  if (grepl("docx", out_format)) {
    return("##### pagebreak")
  } else if (knitr::is_latex_output()) {
    return("\\newpage")
  }
}

@@

Berlin features the most intense Urban Heat Island (UHI) in Germany due to its large extent and development intensity [@kuttler2015], with temperature increases of up to $12~K$ during day-time and $6~K$ on average for night-times [2001-2010, @fenner2014] in urban $vs.$ rural areas.
Consequently, urban green (infrastructure) systems are subjected to increased heat more frequently, potentially affecting their process dynamics - either positively or adversely. 
Their performance and health, however, is closely tied to local energy budgets [@grimmond1996 ; @hertel2019], which in turn are decisive for controlling human wellbeing [e.g. @maras2016], amongst other factors.
Assessing the effect of increased temperatures on green infrastructure, as part of the urban landscape, is therefore instrumental for understanding, and ultimately mitigating, the potential impact of future warming on increasingly urban societies [@norton2015].

Trees, in particular, provide shading as well as transpirative cooling in their vicinity [@gillner2015;@oke1982;@endlicher2016], and therefore can reduce ambient temperatures, infrastructure power-consumption and (human) thermal discomfort  [e.g. @gulyas2006; @akbari2001; @mayer1987; @hoyano1988];
simultaneously, they provide numerous other environmental, cultural and psychological services and/or benefits [see @tzoulas2007 for review].
Further, recent tree growth dynamics as a proxy for on-going and future warming may provide an additional line of evidence to support the growing knowledge base on future climate-vegetation and dynamics [@zhao2016] and may aid in mitigation and adaptation efforts [@pretzsch2017].

Trees and green infrastructure in urban areas show a tendency for enhanced growth rates and/or productivity compared to rural counterparts [@jia2018; @pretzsch2017], yet feature a broad range of effect size ranges and, in some cases, directions specific to species and locality.
 @zhao2016 .. remote sensing.. increased growth rates with higher degree of urbanity for X Chinese cities.
For Berlin, @dahlhausen2018, showed divergent patterns for intermiedate and highly urbanized environments for *Tilia cordata* Mill (using growth increments), the most abundant tree of the city, with increased growth rates in the more intensely developed areas and attributed this to the UHI effect.
<!-- For Berlin, @dahlhausen2018, showed divergent patterns for young and old trees dependent on urban development intensity for *Tilia cordata* Mill, the most abundant tree of the city, but overall reported an increased growth rate which they attributed to the UHI effect. -->
Similarly, @moser-reischl2019a showed positive associations between air temperature and radial growth for two species commonly selected by urban planners (*T. cordata*, *Rubinia pseudoacacia*) in Munich.
By contrast, @gillner2014 highlight decreased growth for *Acer* species  (*A. platanoides* and *pseudoplatanus*), *Platanus x hispanica* and *Quercus rubra* with higher summer temperatures of the preceding year, especially when compounded with drought, in another German metropolis (Dresden). 
Differences in growth trends may result from contrasting intrinsic inter- and intra-specific characteristics, but are indeed affected by other processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc. [@rhoades1999; @quigley2004; @randrup2001; @pauleit2002].
**In the future.. drought, both soil moisture and atmospheric drought.. potential issue for street trees .. [@brune2016]
??**

Conditions affecting tree growth can vary greatly within urban areas or regions, and need to be accounted for when establishing relationships with pertinent drivers, such as the UHI effect.
This typically complicates the extrapolation from individual sampling sites toward predicting effect sizes across entire urban areas and tree stocks, especially for studies reliant on labour-intensive methods limited logistically by sampling effort (reduced sample size, species and spatial coverage).
<!-- While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e. genera and/or species, and space). -->
<!-- However, urban tree growth in urban settings is affected by various processes, which  -->
<!-- However, growth in urban settings is affected by other processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc. [@rhoades1999; @quigley2004; @randrup2001; @pauleit2002]. -->
<!-- While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e. genera and/or species, and space). -->
<!-- This typically hinders the extrapolation from individual sampling sites toward prediction of effect sizes across entire urban areas and tree stocks.  -->
<!-- Therefore, studies reliant on labor-intensive sampling, must carefully address confounding factors to ensure effect sizes are estimated adequately. -->
To complement detailed dendroecological analyses in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individual trees provided by the Berlin Senate Administration (Senatsverwaltung). 
This data set contains information on location, species, trunk diameter (at breast height; $DBH$), and height, amongst other variables. 
In a space-for-time substitution, the absolute growth potential of species can be assessed across the entire urban area, and related to effects of the UHI, while accounting for other location-specific factors, such as street characteristics, development intensity, available soil volume, etc.

[@quigley2004] .. inferred absolute growth potential for species across demographic strata

Note [@gregg2003].. rather impeded growth adjacent to intense urban clusters; but, here focused on intra-city patterns.

@@

# Proposed methods and data requirements

# Preliminary results

@@

# Outlook

@@

<!-- The following line inserts a page break  -->
`r pgbreak()`

@@

`r pgbreak()`
